---
title: "LSSM_documentation"
author: "Edward Gregr"
date: "2025-01-27"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
#Overview

This document summarises the analyses done in R to support the Broughton Local Seaweed Services Model (LSSM) (aka "Models, mesocosms, and field work to assess kelp mitigation of ocean acidification and hypoxia"). 

Relevant aspects include the marine characterisation, the growth of kelp, and the estimation of the effect of kelp forests on the local ocean chemistry. A fundamental aspect of this project is the iterative development of the model. Each aspect is first constructed with using available data and parameters obtained from the literature. The accuracy of the marine characterisation and kelp growth are then improved using local data and a dynamic energy-budget growth model (DEB). The DEB is described below in detail to faciliate the interpretation of the outputs. 

Seasonal growth and OA effects are estimated using a daily temporal resolution, beginning with a 1-dimensional plant model. 

#Marine characterisation

##Temperature and salinity
BATI has 2023 data from 8 moorings in the region. Main ones of interest to this project are B5, which is in Blackfish Sound, and B6, which is in a sheltered location west of Gilford Island, not unlike the Village Island basin. 

The BATI moorings include sensors at different depths:
0.5 m = StarOddi CTD (T/S); 1.5 and 3 m: Hobo (T and Light); Bottom: T (sensor?). These data have been processed by Romina into a workable file of surface T/S with Station ID added (ctd_surface_cond_moorings2023.csv). A second file includes T at depth from the Hobo sensors on these moorings. 

Light data from the Hobos are challenging because the sensors foul up and need cleaning. The data thus need to be cleaned for the deteriorating light levels (this work is pending for Romina). 

MAPP also has 5 moorings in the area. Of interest may be M5 (in QCS) and M4 (at the entrance to Knight Inlet). The MAPP moorings include a StarOddi only, collecting T/S at surface (~ 1m). These data have not yet been processed.

##Alkalinity
Alkalinity is needed to calculate pH. Fortuantely, it has a strong relationship with salinity. BATI moorings have salinity, so can get TA from the following equation (Evans et al. 2015):

  TA = 48.7709*S + 606.23 μmol kg-1 
  
The fit is tight (r2 = 0.94, root mean square error = 17.21 μmol kg-1), and was calculated using only salinity data < 33.6.

##Light simulation
The Daily Light Integral (DLI) is a measure of the total amount of photosynthetically active radiation (PAR) received over a day. We first calculate PAR in units of *(µmol photons m⁻² s⁻²)*. PER HOUR? This was developed with support from ChatGPT. The photoperiod is simulated hourly for a year to match the DEB. DLI is related to PAR as (PARavg X photoperiod / 1e6). So in our case, we just need to daily averages over 1e6 (things coverts micromoles to moles giving DLI in *(mol photons m⁻² day⁻¹*).

Currently can't match Pontier's DLI values without a fudge factor. Check PAR function first. 

##Ambient/reference DIC
Sample data for model development taken from Alaska ferry data. Accessible here:
https://catalogue.hakai.org/erddap/tabledap/HakaiColumbiaFerryResearch.html

Between October 2017 and October 2019, underway surface water measurements were collected from the Alaska Marine Highway System M/V Columbia during 135 service route transits. The vessel sailed south from Ketchikan every Wednesday, arrived in Bellingham on Friday, returned to Ketchikan on Sunday. Measurements of temperature, salinity, dissolved oxygen content, and CO2 partial pressure were made every 2 mins on seawater drawn into the vessel from a ~2 m intake. The CO2 measurement scheme involved the analysis of four gas standards of known CO2 content (150 ppm, 349 ppm, 449 ppm, and 850 ppm; Praxair Distributions Inc), 12 analyses atmospheric CO2, and 240 seawater CO2 measurements in a 8.5 hour period. The atmospheric and seawater CO2 analyses were run in a sequence of 3 atmospheric measurements and 60 seawater measurements that was repeated 4 times between standardization. 
Of interest here is seawater pCO2, and this is measured in 



#Kelp growth
##Literature-based growth
Weigel and Pfister (2021) sampled kelp at peak biomass (June to Sept). They found:
Avg blade length     = 142+/-8.5 cm
Avg dry mass         = 4.13+/-0.27 g
Scaled C fixing rate = 2.35+/- kg C / m2.yr. This is NET primary production, after accounting for the proportion C lost as respiration (0.27+/-0.08), of which, based on other studies (REF), 0.16 is released as DOC.

This C fixing rate is based on an estimate of daylight hours (Apr-Sep) = 2617. 
This is extended to the fixing rate of a KELP FOREST using :
  mean no. of plants per m2    = 7.66+/-0.86
  mean no. of blades per indiv = 41.5+/-2.95
They note that earlier density measurements are lower:  2.25 (Foreman 1984), 3.7 (Barns and Kalvass 1993), and 4.57 indiv per m2 (Stekoll et al. 2006).

Translating a fixing rate of kgC/m2/yr to a plant:
  2.35/7.66 = 0.30 kg C / yr / plant 

As a realism check, convert the carbon fixed to wet weight. Using available conversion factors:
  dw   = 0.13 * ww (Nereo, from Bullen)
  kgC  = 0.25 kg dw

Leads to an estimate of 9.23 (0.30/0.13/0.25) kgWW / yr / plant. Not unreasonable. 

This can  be scaled to daily or monthly growth, by disaggregating total annual daylight hours (Weigel and Pfister 2021). From this we could also back calculate: 
kg C / frond, which would allow comparisons to the results of the DEB and may help parameterise scaling of the DEB to a plant. 

##Dynamic energy budget growth model
The Dynamic energy budget (DEB) developed by XXX was calibrated for nereocystis by Barbosa. 

The dynamics of the DEB are represented by three state variables: 
  Mv (MolV)                  = the structural mass, 
  meN = MeN/Mv (molN molV-1) = the nitrogen reserve density, 
  meC = MeC/Mv (molC molV-1) = and the carbon reserve density. 

MeC and MeN are the masses of the C and N reserves. 

The DEB takes a table of [temperature, nutrients, dissolved inorganic carbon (DIC), and light] as an input:

T =  K,          temperature
N =  molN/L,     dissolved inorganic nitrogen concentration
C =  molC/L,     dissolved inorganic carbon concentration
I = molÎ³/m^2/d, photon concentration (= irradiance)

The model returns a table of results for each time step. This table includes a few dozen parameters, with names that follow a standardized approach [see DEB book], summarised below: 

*j* represents a flux rate
*E* represents a reserve (or energy) unit, for either N or C
*M* represents molar mass 
*V* represents the volume of the structure (i.e., frond) in L^3
*G* is the growth pool for a flux
*R* is the reproduction pool for a flux

*y* represents yield factors that couple flux processes 
*t* as a suffix on the variable names, it represents the time step of the variable

The following parameters are (assumed to be) of use to assessing growth effects on ocean chemistry:

Assimilation variable typical pathways:
  j_NA = molN/molV/h,   uptake rate of N
  j_CI = molDIC/molV/h, uptake rate of DIC
  j_I  = specific relaxation rate
  j_CA = molC/molV/h, DIC assimilation rate
  j_Ot  = O2 production
  j_Ct  = Carbon (where?)

Allocation to growth:
  j_EN_G = specific flux of N to growth SU (after maintenance)
  j_EC_G = specific flux of C to growth SU (after maintenance)
  j_G    = gross specific growth rate

Total weight and length are closely correlated, as length is a species-specific function of biomass.
  Wt        = g
  L_allomt = cm, physical length

Integration of results into the state variables for next steps: 
  M_V  = M_V + dMVdt * dt       The volume of the blade in terms of moles;
  m_EN = m_EN + dmENdt * dt     The mass of the nitrogen reserve;
  m_EC = m_EC + dmECdt * dt;    The mass of the carbon reserve
  
Molar weights (g/mol) are used for converting moles (the mass unit in the DEB) to grams. These are listed in Lavaud et al.'s Table 2. The values for carbon and nitrogen reserves are higher than their atomic masses as they reflect the average molar mass for the compounds typically found in the reserve pool of kelps. Minerals and inorganics can also make up 15-25% of kelp's dry weight (DEB's typically estimate dry weight).
  Molar weight of structure wV      27.51
  Molar weight of N reserve wEN     17
  Molar weight of C reserve wEC     30

#Ocean chemistry
A critical aspect of estimating the change in pH attributable to kelp.
With the data described above, we used seacarb() to estimate the change in pH attributable to kelp.



### Other BATI ocean data
T/S profiles were collected during kelp sampling. Likely only relevant to water column investigations. 
Environmental button loggers (T only) distributed around the region and collecting data since 2022. Unclear on the number and locations, but should be useful for spatial variance in T. 

## Field data 
Field observations of nereo helped parameterize growth and distribution

## Spatial data
We built clusters to estimate the abundance and spatial distribution of nereocystis.

# Methods

## BUilding an environmental conditions layer
The DEB model takes a table of [light, nutrients, temperature, and DIC] as input. 

Currently, this is built with constant but randomized light, nutrients and DIC. Temperature is taken from the BATI5 sensor, with the intent of comparing the pH results to BATI6, thereby comparing a QCS reference (BATI5) with and estuarine site (BATI6).

```{r envData, echo=FALSE, fig.pos='h', fig.cap="Preliminary Drivers of the Dynamic Energy Budget Model - BATI5", fig.align='center'}

PlotInputs( grow_dat, "May to September Environmental Conditions")


```
# Results 
Environmental layers are passed through a compilied DEB, leading to an assortment of outputs. 

Relevant calculations include:

Total C assimilated = sum of carbon assimilation per time step 
                    = sum of assimilation rate * volume of kelp
                    = sum( j_CAt [molC/molV/h] * M_Vt [molV] )

This total partitions to growth and maintenance. 

Total C fixed for growth = sum of carbon used for growth per time step
                         = sum( j_EC_G [molC/molV/h] * M_Vt [molV] )

The difference in these values would be the C used for maintenance, which is mostly released as respiration. 


Next step is to figure out how to estimate change in DIC from the output data, along with a measure of total plant size. This would include using below-bulb stype diameter to estimate stipe mass, plus some number of fronds to multiply by the growth value. (These fronds will need to be shed at some rate ... see Canvin et al. 2024).

Can we assume that the growth of the entire plant follows the growth rate of a blade? Not exactly according to Pontier et al. 2024.



## Values from elsewhere

ChatGPT says:
- A mature bull kelp (*Nereocystis luetkeana*) plant can have 30 - 50 blades
- The stipe usually makes up around 20-40% of the total biomass, depending on age and growth stage, season, and environmental factors. 
- A new sporophyte is typically a few mm to cm in length, and a few to 10s of mg (dry weight).


#References

Evans et al. 2015. Tracking Ocean Acidification in an Alaskan Shellfish Hatchery, PLoS One, 10, e0130384,
https://doi.org/10.1371/journal.pone.0130384.



Fin.